\par{

\subsection{Partial Dependence Function}

It is useful to have a coherent method of measuring the importance a specific variable $X_j$ has on the output $Y$. 
Since partial dependence plots usually only depict the relationship between two variables, we follow Klusowski (2019) by introducing a conditional partial dependence function. 
Recall that we assume $Y_i = m(\mathbf{X}_i) + \epsilon_i$. 
The idea behind the conditional partial dependence function is to look at the influence of one specific $X_j$ while ignoring the others. 
For this let 

\begin{align}
    \bar{F}_j(x_j; \mathbf{t}) = E [Y \ | \ \mathbf{X} \in \mathbf{t}, X_j = x_j] \label{pardepfunc_f}
\end{align}

where $\bar{F}_j(x_j; \mathbf{t})$ is the partial dependence function for variable $j$ conditional on being in node $\mathbf{t}$. 
One can also think of () as a solution to a least squares approximation of $m(\mathbf{X})$ as a function of only $X_j$ within node $\mathbf{t}$.
For later purposes it will be beneficial to additionally define the mean centered partial dependence function 

\begin{align}
    \bar{G}_j(x_j; \mathbf{t}) &= \bar{F}_j(x_j; \mathbf{t}) - E [Y \ | \ \mathbf{X} \in \mathbf{t}] \nonumber \\
    &= E [Y \ | \ \mathbf{X} \in \mathbf{t}, X_j = x_j] - E [Y \ | \ \mathbf{X} \in \mathbf{t}] \label{pardepfunc_g}
\end{align}

which is the partial dependence function demeaned by the average of all observation within the respective node. \\

This concept now allows us to reformulate our definition of strong and weak variables.
Since weak variables $\{X_j : j \notin \mathcal{S}\}$ are independent of $Y$, by law of iterated expectations, (\ref{pardepfunc_f}) is equal to $E [Y \ | \ \mathbf{X} \in \mathbf{t}]$.
This in turn implies that the demeaned partial dependence function $\bar{G}_j(x_j; \mathbf{t})$ is zero for all weak variables.
In contrast, strong variables, have an effect on $Y$ and thus the demeaned partial dependence function is not necessarily 0. 
Note however, that here we concern ourselves with the population parameters. 
In a finite sample analysis, it is indeed the case that because of the error term, we can have $\bar{G}_j(x_j; \mathbf{t}) \neq 0$ even for weak variables. 
This will fact play an important part in the later analysis when constructing a decision tree. 